---
title: "Geo_Data_Sci_R"
author: "Wyclife Agumba Oluoch"
date: "1/22/2021"
output: html_document
---

### Loading Libraries

Loading the necessary packages to use in the codes which will be helpful in 
analyzing geographic data in R.

```{r packages, include=FALSE}
library(sf)
library(tidyverse)
```

The data which I will use in the current work can be [directly downloaded](https://opendata.arcgis.com/datasets/f99ce43936d74f718e92a37a560ad875_0.zip), extracted, and saved in the subfolder called *shp* inside the .Rmd file folder.

The next step is to read in the data using `st_read()`. I will call the created object honeybee

```{r loading, echo=FALSE, message=FALSE, warning=FALSE}
honeybee <- st_read('shp/HoneyBeePermits2017.shp')

```

There is a lot of information which we get by loading the data. We can extract such information individually, for example:

```{r geometry}
str(honeybee$geometry) 
```

This is telling us that we are working with a point feature. Such a cool information to have. Now we may need to know the coordinate reference system of our data

```{r coordinaes}
st_crs(honeybee)
```

This is returning a super helppful information. We can also have the information within the text as `r st_crs(honeybee)`. I still have no idea how the output will look like. Let us see by knitting the markdown. Everything thrown to text.







